---
title: "Creating Sequences with TraMineR"
author: "Chloe Hutton"
date: "2024-08-12"
output: html_document
---

# 02 Creating sequences with TraMineR

`TraMineR` was used to turn the csv file of patient trajectories into a sequence object.

First, the necessary libraries were imported.

```{r}
# Load required libraries
library(TraMineR)  # for creating the sequence object and plots
library(RColorBrewer)  # for colour palettes
```

Next, the working directory was set (update if necessary).

```{r setup}
knitr::opts_knit$set(root.dir = "~/treatment_pathway_exploration")
```

Next, the data file directory was created as a variable.

```{r}
# patient trajectories
data_file <- "./data/distinct_GLP1_sequences.csv"
```

For `TraMineR`, you can specify the following:

-   `alphabet` are all the values existing in the trajectory file currently

-   `palette` allows you to specify the colours to use in the sequence plots

-   `labels` specifies the descriptions for every state (needs to be in the same order to correspond with the *alphabet*)

-   `scodes` are short-hand versions of the labels handy if you need to use shorter descriptions

```{r}
alphabet <- c(0, 1, 2, 3, 4, 5, 6, 7)
palette <- brewer.pal(length(alphabet), 'Set2')
labels <- c("No GLP-1 RA or Other Anti-Obesity Medication",
            "Semaglutide",
            "Liraglutide",
            "Dulaglutide",
            "Tirzepatide",
            "Exenatide",
            "Other Anti-Obesity Medication",
            "GLP-1 RA & Other Anti-Obesity Medication")
scodes <- c("No medication",
            "Semaglutide",
            "Liraglutide",
            "Dulaglutide",
            "Tirzepatide",
            "Exenatide",
            "Other",
            "GLP-1 RA & Other")
```

## Loading the data

The trajectories are loaded.

```{r}
# Load the data
# check names is FALSE to stop column names being reformatted
trajact <- read.table(data_file, header = TRUE, sep = ",", row.names = 1, check.names=FALSE)
```

The trajectories were previewed to check they loaded as expected.

```{r}
head(trajact)
```

## Creating the sequence object

The variables above were used to define the sequence state object.

```{r}
seqact <- seqdef(trajact, cpal = palette,
                 alphabet = alphabet,
                 states = scodes,
                 labels = labels,
                 xstep = round(ncol(trajact) / 43, 0),
                 tick.last = TRUE)
```

Again, we can summarise the result.

```{r}
summary(seqact)
```

## Viewing the entropy and state distribution

Two key plots of the full cohort can be accessed as functions from the file `full_cohort_analysis.r` , which we can load into this notebook to use.

```{r}
# load plotting functions for full cohort analysis
source("./code/full_cohort_analysis.r")
```

To save down a copy of the entropy over time, use the `create_entropy_plot` function. You can manually specify the file path and file name if you do not wish for it to be saved under the default option, which is saved in results as *Entropy_Plot.png*.

```{r}
create_entropy_plot(seqact, plot_file = "./results/Example_entropy_plot.png")
```

For the state distribution plot over time, use the `create_dist_plot` function. For this, you will need to specify the labels and palette.

```{r}
create_dist_plot(seqact, labels, palette, plot_file = "./results/Example_distribution_plot.png")
```

## Viewing the entropy for GLP-1 RA medications by type and pooled

The entropy plot can be viewed with two lines if required. To show entropy over time for both distinct and pooled GLP-1 RAs, we will load in a new sequence object.

```{r}
# file for pooled GLP-1 RAs
data_file_pooled <- "./data/GLP1_sequences.csv"

# create alphabet, palette, labels and scodes for pooled version
alphabet_pooled <- c(0, 1, 2, 3)
palette_pooled <- brewer.pal(length(alphabet_pooled), 'Set2')
labels_pooled <- c("No GLP-1 RA or Other Anti-Obesity Medication",
                   "GLP-1 RA",
                   "Other Anti-Obesity Medication",
                   "GLP-1 RA & Other Anti-Obesity Medication")
scodes_pooled <- c("No medication",
                   "GLP-1 RA",
                   "Other",
                   "GLP-1 RA & Other")

# Load the data
trajact_pooled <- read.table(data_file_pooled, header = TRUE, sep = ",", row.names = 1, check.names=FALSE)

# create pooled GLP-1 RAs sequence object
seqact_pooled <- seqdef(trajact_pooled, cpal = palette_pooled,
                            alphabet = alphabet_pooled,
                            states = scodes_pooled,
                            labels = labels_pooled)
```

Once loaded, you can create the entropy plot.

```{r}
create_entropy_plot(seqact, seqact2 = seqact_pooled, plot_file = "./results/Example_two_entropies_plot.png")
```

## Working out the random entropy of the trajectories

To provide a benchmark for the clustering of future models, the random entropy score for the full and excluding pre-index date, was calculated.

First, we will load the file containing the random entropy function.

```{r}
source("./code/random_entropy.r")
```

```{r}
# Setup parallel processing
num_cores <- detectCores() - 1
cl <- makeCluster(num_cores)
registerDoParallel(cl)

# Set seed for reproducibility
set.seed(101)
print("Starting random entropy measures...")
start_rand_entropy <- Sys.time()

# Compute random entropy scores in parallel
# project used 100x but this example does 5 for demo
rand_entropy_all <- foreach(i = 1:5, .combine = c, .packages = 'entropy') %dopar% {
    rand_entropy_measure(seqact_pooled)
}

rand_entropy_rhs <- foreach(i = 1:5, .combine = c, .packages = 'entropy') %dopar% {
    rand_entropy_measure(seqact_pooled, rhs = TRUE)
}

end_rand_entropy <- Sys.time()
rand_ent_time <- round(difftime(end_rand_entropy, start_rand_entropy, units = "secs"))

# Cleanup
stopCluster(cl)

# Print the mean result to 3 decimal points
print(round(mean(rand_entropy_all)), 3)
print(round(mean(rand_entropy_rhs)), 3)
sprintf("Total random entropy time at 100 replications: %s seconds", rand_ent_time)
```

The results can be saved down to a csv file.

```{r}
# Custom column headings
column1_name <- "Random_Entropy_all"
column2_name <- "Random_Entropy_After_Index"

# Create a data frame from the vectors
df <- data.frame(rand_entropy_all, rand_entropy_rhs)

# Set the column names
colnames(df) <- c(column1_name, column2_name)

# Specify the path to save the CSV file
file_path <- "./data/EXAMPLE_random_entropy_pooled_stratified.csv"

# Save the data frame to a CSV file
write.csv(df, file_path, row.names = FALSE)
```
