---
title: "Creating a time series clustering model"
author: "Chloe Hutton"
date: "2024-08-20"
output: html_notebook
---

# 03 Creating a time series clustering model

As before, we will ensure that the working directory is correct.

```{r setup}
knitr::opts_knit$set(root.dir = "~/treatment_pathway_exploration")
```

## Loading in the sequence objects and TrajectoryAnalysis object

To make it easier, we will load in the `patient_trajectories.r` file which will create and load the following:

-   `seqact_all` all patient trajectories for the cohort with a state for every GLP-1 RA type

-   `seqact_pooled_all` all patient trajectories for the cohort with one state for all GLP-1 RAs

-   `create_stratified_cohort` is a function to create a sample of the full cohort

-   `TrajectoryAnalysis` is an `R6` class for creating the time series clustering model

Additionally, there are separate variables for `palette`, `scodes`, `labels` and `alphabet` - for GLP-1 RAs as one state, add \*\_pooled\* to the end (e.g. `palette_pooled`)

```{r}
source("./code/patient_trajectories.r")
```

The `create_stratified_cohort` function can be used to sample the dataset stratifying by age, gender, year of index date and other anti-obesity medication use. The default sample size is 10%, but you can adjust this with `sample_size`, specifying the sample size as a decimal.

```{r}
# additional patient information for stratifying sample
pat_details_file <- "./data/pat_level_info.csv"

# run function to create a stratified sample
stratified_seqact_pooled <- create_stratified_cohort(seqact_pooled_all, pat_details_file = pat_details_file, sample_size = 0.1)
```

```{r}
# check summary information is a 10% sample of full cohort
summary(stratified_seqact_pooled)
```

Example with distinct GLP-1 RA types.

```{r}
# run function to create a stratified sample with GLP-1 RA types
stratified_seqact <- create_stratified_cohort(seqact_all, pat_details_file = pat_details_file, sample_size = 0.1)
```

```{r}
# check summary information
summary(stratified_seqact)
```

## Using the `TrajectoryAnalysis` class to create a model

To create a model, we will use the pooled GLP-1 RA stratified sample, using a transition rate substitution cost matrix which varies for every time unit, to be used with dynamic hamming distance (*DHD*).

For this, we need to specify:

-   `time_varies` needs to be **TRUE** to ensure the substitution cost matrix varies every time unit.

-   `seqdist_method` needs to be **DHD** for dynamic hamming distance

Other parameters:

-   `pi_dissim` indicates whether you wish for the distance matrix to be created on the post-index date period only or not. It defaults to **FALSE** but in this instance, we will switch it to **TRUE** as there is little heterogeneity in the pre-index date period.

-   `name` allows you to provide the model with a suitable name.

-   `rand_entropy` and `rand_entropy_all` are two optional parameters if you wish for the final results table to include the random entropy scores for comparison.

-   `trace_flag` is set to **TRUE** to print out extra details while the methods in the class are run. This can be changed to **FALSE** to reduce the print output in the console.

-   `heatmap_flag` is an option to save down a sequence index plot during the `run_analysis` method. For now, we will leave this as **FALSE**.

```{r}
model <- TrajectoryAnalysis$new(stratified_seqact_pooled, palette_pooled,
                                alphabet_pooled, labels_pooled, scodes_pooled,
                                name = "Model_A_stratified", con_val = 2,
                                cost_method = "TRATE", trace_flag = TRUE,
                                heatmap_flag = FALSE, seqdist_method = "DHD",
                                time_varies = TRUE,
                                cluster_method = "ward.D2",
                                rand_entropy = 0.33,
                                rand_entropy_rhs = 0.628,
                                pi_dissim = TRUE)
```

One a model is created, the method `run_analysis` is used to run the model.

```{r}
# run the model
model$run_analysis()
```

## Reviewing a model's results

There is an option to review the results.

```{r}
model$results
```

There are other methods in the class which can be used to review the model.

-   `cluster_heatmap` provides a sequence index plot of the patient trajectories with partitions to indicate the clusters. There is a parameter, `k`, to determine the number of clusters to show.

-   `dend_heatmap` is similar to the above, but shows the dendrogram on the left side of the plot for further context/ However, it does not have y-axis tick labels.

-   `create_dendrogram` saves down the dendrogram on its own.

```{r}
model$cluster_heatmap(k = 2)
```

```{r}
model$dend_heatmap(k = 2)
```

```{r}
model$create_dendrogram()
```

## Saving down a model to review later

If you have run a model on the full cohort and wish to save time in the future when you wish to review it again, you can save down a model.

```{r}
# save model for future use
saveRDS(model, file = sprintf("./models/%s.rds", model$name))
```
