---
title: "Stratified sample model iterations"
author: "Chloe Hutton"
date: "2024-08-20"
output: html_notebook
---

# 04 Stratified sample model iterations

To find a suitable model for the cohort, a stratified sample was created to iterate through different model parameters.

## Loading in the sequence object and creating a stratified sample.

First, we will set the working directory and then load in the sequence objects, `TrajectoryAnalysis` class and `create_stratified_cohort` function.

```{r setup}
knitr::opts_knit$set(root.dir = "~/treatment_pathway_exploration")
```

```{r}
source("./code/patient_trajectories.r")
```

```{r}
# additional patient information for stratifying sample
pat_details_file <- "./data/pat_level_info.csv"

# run function to create a stratified sample - very small sample for demo!
stratified_seqact_pooled <- create_stratified_cohort(seqact_pooled_all, pat_details_file = pat_details_file, sample_size = 0.01)
```

## Defining the parameters

Next, we will create a series of variables to consider for the model parameters.

```{r}
# Define the parameter grid variables
substitution_costs <- seq(0.5, 10, by = 0.5)
sub_method_param <- c("CONSTANT", "TRATE", "INDELSLOG", "FUTURE")
time_varies_param <- c(TRUE, FALSE)
lag_param <- c(1, 7, 14, 21, 28, 54, 90, 104)
indel_param <- seq(0.5, 3, by = 0.5)
dist_method_param <- c("HAM", "DHD", "OM", "OMspell", "CHI2", "TWED")
tpow_params <- seq(0.1, 2, by = 0.5)
expcost_params <- seq(0.1, 1, by = 0.1)
h_params <- seq(0.1, 1, by = 0.1)
nu_params <- seq(0.1, 1, by = 0.1)

# Define 28-day intervals
breaks_28 <- list()
for (i in seq(1, 365, by=28)) {
  end <- min(i + 27, 365)
  breaks_28 <- append(breaks_28, list(c(i, end)))
}

# Define 30-day intervals
breaks_30 <- list()
for (i in seq(1, 365, by=30)) {
  end <- min(i + 29, 365)
  breaks_30 <- append(breaks_30, list(c(i, end)))
}
# Combine the two sets of intervals
breaks_combined <- c(breaks_28, breaks_30)
#breaks <- list(c(1:28), c(1,30), c(1,31), c(28, 54), c())
step_params <- seq(0.1, 1, by = 0.1)
overlap_params <- c(TRUE, FALSE)
#range_1_param <- list(NA, 1:28, 1:54, 1:90, 1:104)
range_1_param <- IRanges(start=c(0, 1, 1, 1, 1), end=c(0, 28, 54, 90, 104))
range_1_val_param <- seq(1, 20, by = 1)
```

## Create a results table and file for saving down the results

```{r}
# Initialize data frame to store all parameter models and their results
results_table <- data.frame(ModelName = character(),
                            TimeTaken = numeric(),
                            DistMatrixTime = numeric(),
                            ClusteringTime = numeric(),
                            OLOTime = numeric(),
                            ASWw = numeric(),
                            ASWw_num_clusters = numeric(),
                            CHsq_value = numeric(),
                            CHsq_clusters = numeric(),
                            Entropy = numeric(),
                            EntropyRHS = numeric(),
                            EntropyRandom = numeric(),
                            EntropyRandomRHS = numeric(),
                            SeqDistMethod = numeric(),
                            ClustMethod = character(),
                            SubMatrixMethod = character(),
                            ConstantValue = numeric(),
                            Range1_range = character(),
                            Range1_val = numeric(),
							              Indel = numeric(),
                            NPatients = numeric(),
                            stringsAsFactors = FALSE  # Avoid automatic factor conversion
							              )

# create the results file
results_file = "./results/gridsearchresults-sample.csv"

# write the column headings to the file
write.csv(results_table, results_file, row.names = FALSE)

```

## Iterate over all the parameters

```{r}
# Grid search - iterate through options
for (dist_method in dist_method_param) {
  for (sub_cost in substitution_costs) {
    for (sub_method in sub_method_param) {
      for (lag in lag_param) {
        for (indel in indel_param) {
          for (time_varies in time_varies_param) {
            # Skip invalid combinations
            if ((time_varies == TRUE && !(dist_method %in% c("DHD"))) |
                (dist_method == "DHD" && isFALSE(time_varies)) |
                (dist_method == "DHD" && sub_method == "FUTURE") |
                (dist_method %in% c("DHD", "HAM", "CHI2") && indel > 0.5) |
                (sub_method == "INDELSLOG" && indel > 0.5)|
                (!(sub_method %in% c("CONSTANT", "TRATE")) && sub_cost > 0.5)) {
              next
            }
            if (sub_method == "INDELSLOG") {
              indel <- "auto"
            }
            # Skip to avoid iterating over lags for methods not applicable
            if (lag > 1 && !(sub_method %in% c("TRATE", "FUTURE"))) {
              next
            }
            # if distance matrix is OMspell
            if (dist_method == "OMspell") {
              for (tpow in tpow_params) {
                for (expcost in expcost_params) {
                  model_name <- sprintf("OMspell-%stpow-%sexpcost-%sindel-%scval", tpow, expcost, indel, sub_cost)
                  model <- TrajectoryAnalysis$new(stratified_seqact_pooled, palette_pooled,
                                                  alphabet_pooled, labels_pooled,
                                                  scodes_pooled, name = model_name,
                                                  con_val = sub_cost,
                                                  cost_method = sub_method,
                                                  seqdist_indel = indel,
                                                  trace_flag = trace_flag,
                                                  heatmap_flag = heatmap_flag,
                                                  tpow = tpow,
                                                  seqdist_method = dist_method,
                                                  cluster_method = "ward.D2",
                                                  expcost = expcost,
                                                  rand_entropy = rand_entropy_all,
                                                  rand_entropy_rhs = rand_entropy_rhs,
                                                  pi_dissim = TRUE)
                }
              }
              # if TWED
            } else if (dist_method == "TWED") {
              for (h in h_params) {
                for (nu in nu_params) {
                  model_name <- sprintf("TWED-%sh-%snu-%sindel-%scval", h, nu,
                                        indel, sub_cost)
                  model <- TrajectoryAnalysis$new(stratified_seqact_pooled, palette_pooled,
                                                  alphabet_pooled, labels_pooled,
                                                  scodes_pooled, name = model_name,
                                                  con_val = sub_cost,
                                                  cost_method = sub_method,
                                                  trace_flag = trace_flag,
                                                  heatmap_flag = heatmap_flag,
                                                  h = h, seqdist_indel = indel,
                                                  seqdist_method = dist_method,
                                                  cluster_method = "ward.D2",
                                                  nu = nu,
                                                  rand_entropy = rand_entropy_all, rand_entropy_rhs = rand_entropy_rhs,
                                                  pi_dissim = TRUE)
                }
              }
              # if CHI2
            } else if (dist_method == "CHI2") {
              
              for (step in step_params) {
                model_name <- sprintf("CHI2-%sbreak-%sstep-%sindel-%scval",
                                      break_param, step, indel, sub_cost)
                model <- TrajectoryAnalysis$new(stratified_seqact_pooled, palette_pooled,
                                                alphabet_pooled, labels_pooled,
                                                scodes_pooled, name = model_name,
                                                con_val = sub_cost,
                                                cost_method = sub_method,
                                                trace_flag = trace_flag,
                                                heatmap_flag = heatmap_flag,
                                                breaks = breaks_combined,
                                                seqdist_method = dist_method,
                                                cluster_method = "ward.D2",
                                                step = step,
                                                rand_entropy = rand_entropy_all, rand_entropy_rhs = rand_entropy_rhs,
                                                pi_dissim = TRUE)
                
              }
              # if DHD and time varies for substitution cost matrix
            } else if (dist_method == "DHD" && isTRUE(time_varies)) {
              print(range_1_param)
              for (i in 1:length(range_1_param)) {
                print(end(range_1_param[i]))
                if (end(range_1_param[i]) == 0) {
                  # Skip constant
                  if (sub_method == "CONSTANT") {
                    next
                  } else {
                    model_name <- sprintf("DHD-%ssubmethod-%sindel-%scval", sub_method, indel, sub_cost)
                    model <- TrajectoryAnalysis$new(stratified_seqact_pooled, palette_pooled,
                                                    alphabet_pooled, labels_pooled,
                                                    scodes_pooled,
                                                    name = model_name,
                                                    con_val = sub_cost,
                                                    cost_method = sub_method,
                                                    trace_flag = trace_flag,
                                                    heatmap_flag = heatmap_flag,
                                                    seqdist_method = dist_method,
                                                    cluster_method = "ward.D2",
                                                    time_varies = TRUE,
                                                    rand_entropy = rand_entropy_all,
                                                    rand_entropy_rhs = rand_entropy_rhs,
                                                    pi_dissim = TRUE)
                  }
                  # all the rest
                } else {
                  for (rval in range_1_val_param) {
                    model_name <- sprintf("DHD-%srange1-%srangeval-%sindel-%scval",
                                          range, rval, indel, sub_cost)
                    rng1 <- start(range_1_param[i]):end(range_1_param[i])
                    model <- TrajectoryAnalysis$new(stratified_seqact_pooled, palette_pooled,
                                                    alphabet_pooled, labels_pooled,
                                                    scodes_pooled,
                                                    name = model_name,
                                                    con_val = sub_cost,
                                                    cost_method = sub_method,
                                                    trace_flag = trace_flag,
                                                    heatmap_flag = heatmap_flag,
                                                    range_1 = rng1,
                                                    range_1_val = rval,
                                                    seqdist_method = dist_method,
                                                    cluster_method = "ward.D2",
                                                    time_varies = TRUE,
                                                    rand_entropy = rand_entropy_all,
                                                    rand_entropy_rhs = rand_entropy_rhs,
                                                    pi_dissim = TRUE)
                  }
                }
              }
            } else {
              # Handle other methods (e.g., HAM) here as needed
              model_name <- sprintf("%s-%s-%sindel-%scval",
                                    sub_method, dist_method, indel, sub_cost)
              model <- TrajectoryAnalysis$new(stratified_seqact_pooled, palette_pooled,
                                              alphabet_pooled, labels_pooled,
                                              scodes_pooled, name = model_name,
                                              con_val = sub_cost,
                                              cost_method = sub_method,
                                              trace_flag = trace_flag,
                                              heatmap_flag = heatmap_flag,
                                              seqdist_indel = indel,
                                              seqdist_method = dist_method,
                                              cluster_method = "ward.D2",
                                              rand_entropy = rand_entropy_all,
                                              rand_entropy_rhs = rand_entropy_rhs,
                                              pi_dissim = TRUE)
              
            }
            model$run_analysis()
            print(model$results)
            
            write.table(model$results, results_file,
                          append = TRUE,
                          sep = ",",
                          col.names = FALSE,
                          row.names = FALSE,
                          quote = FALSE)
            # Store the current parameters and quality in the results table
            results_table <- rbind(results_table, model$results)
          }
        }
      }
    }
  }
}

```

## Review the results

You can print the results at the end of the run or view the saved csv file.

```{r}
# Print the results table
print(results_table)
```
