---
title: "Analysing models further"
author: "Chloe Hutton"
date: "2024-08-20"
output: html_notebook
---

# 05 Analysing models further

Once a model or models have been selected for further analysis, you can use some functions in the `model_analysis.r` file to analyse it further.

First, we will set the directory and then load in the functions.

```{r setup}
knitr::opts_knit$set(root.dir = "~/treatment_pathway_exploration")
```

```{r}
# load model analysis functions
source("./code/model_analysis.r")
```

```{r}
# load sequence objects
source("./code/patient_trajectories.r")
```

For this example, we will load the example model saved down from notebook 03, or you can update to another model if you have saved one.

```{r}
model_A <- readRDS("./models/model_A_stratified.rds")
```

We will obtain the stratified sample used as well for GLP-1 RA type in case we want to use it in some of the visuals. We will filter to the patients in the stratified sample for pooled, using the row names.

```{r}
# run function to create a stratified sample with GLP-1 RA types
stratified_seqact <- seqact_all[row.names(model_A$seqact),]

summary(stratified_seqact)
```

We can create an intertia plot using the `create_inertia_plot` function from `model_analysis.r`

```{r}
# create inertia plot using height of dendrogram
create_inertia_plot(model_A, elbow_point = NA)
```

We can create labelled clusters with the `label_clusters` function.

```{r}
# create labelled clusters for plotting
model_A_clusters <- label_clusters(model_A, k = 2)
```

You can preview the distribution of patients between clusters.

```{r}
table(model_A_clusters)
```

The state distribution plot can be produced again but this time stratified by cluster.

```{r}
# create distribution plots split by cluster
create_dist_plot(seqact_all[row.names(model_A$seqact),], scodes, palette, clusters = model_A_clusters, model_name = model_A$name)
```

The representative sequences per cluster can be output as well to an image file.

```{r}
# GLP-1 RA state sequence representative plot
create_rep_plot(model_A$seqact, model_A, labels_pooled, palette_pooled, clusters = model_A_clusters)
```

As there is little heterogeneity in the pre-index date period, we will plot the mean time spent in every state for index date onwards instead.

```{r}
create_mean_time_plot(stratified_seqact, model_A, labels, palette, clusters = model_A_clusters)
```

To analyse the clusters further, such as patient characteristics, we can save the patient IDs (row names) and clusters down for later.

```{r}
# save down clusters by PatId into a csv in results folder
save_clusters_id(model_A, k = 2)
```
