---
title: "Producing summary tables from patient characteristics and treatment durations"
author: "Chloe Hutton"
date: "2024-08-20"
output: html_notebook
---

# 06 Producing summary tables from patient characteristics and treatment durations

First, we will load the relevant libraries.

```{r}
# load libraries
library(readr)
library(dplyr)
library(pROC)
library(broom)
library(gtsummary)
```

Next, we will set the working directory.

```{r setup}
knitr::opts_knit$set(root.dir = "~/treatment_pathway_exploration")
```

The location for patient level information needs to be identified next.

```{r}
# file directory
pat_details_file <- "./data/pat_level_info.csv"
```

We will load in the data but filter to the columns we need.

```{r}
# read the data
pat_details <- read.table(pat_details_file, header = TRUE, sep = ",", row.names = 1, check.names = FALSE)

# refine to only the required columns
cols_to_use <- c("Gender:Description",
                 "Age at Index Date:Description",
                 "DULAGLUTIDE:Event Occurrence",
                 "DULAGLUTIDE:Sum:Days Supply",
                 "EXENATIDE  (BYDUREON, BYETTA):Event Occurrence",
                 "EXENATIDE  (BYDUREON, BYETTA):Sum:Days Supply",
                 "LIRAGLUTIDE (SAXENDA):Event Occurrence",
                 "LIRAGLUTIDE (SAXENDA):Sum:Days Supply",
                 "LIRAGLUTIDE (VICTOZA):Event Occurrence",
                 "LIRAGLUTIDE (VICTOZA):Sum:Days Supply",
                 "LIXISENATIDE ( ADLYXIN):Event Occurrence",
                 "LIXISENATIDE ( ADLYXIN):Sum:Days Supply",
                 "Qsymia, Contrave, Xenical, Alli:Event Occurrence Before Index Date",
                 "Qsymia, Contrave, Xenical, Alli:Event Occurrence on or After Index Date",
                 "Qsymia, Contrave, Xenical, Alli:Sum:Days Supply",
                 "SEMAGLUTIDE (OZEMPIC):Event Occurrence",
                 "SEMAGLUTIDE (OZEMPIC):Sum:Days Supply",
                 "SEMAGLUTIDE (WEGOVY):Event Occurrence",
                 "SEMAGLUTIDE (WEGOVY):Sum:Days Supply",
                 "TIRZEPATIDE (MOUNJARO):Event Occurrence",
                 "TIRZEPATIDE (MOUNJARO):Sum:Days Supply",
                 "ALBIGLUTIDE (TANZEUM):Event Occurrence",
                 "ALBIGLUTIDE (TANZEUM):Sum:Days Supply")             

# columns to not include in summary table
summary_data <- pat_details[, cols_to_use]
# preview
head(summary_data)
```

Next, we will improve the readability of the columns chosen.

```{r}
colnames(summary_data) <- c("Gender",
                           "Age at Index Date",
                           "Dulaglutide (Trulicity)",
                           "Dulaglutide (Trulicity) Duration",
                           "Exenatide (Bydureon, Byetta)",
                           "Exenatide (Bydureon, Byetta) Duration",
                           "Liraglutide (Saxenda)",
                           "Liraglutide (Saxenda) Duration",
                           "Liraglutide (Victoza)",
                           "Liraglutide (Victoza) Duration",
                           "Lixisenatide (Adlyxin)",
                           "Lixisenatide (Adlyxin) Duration",
                           "Qsymia, Contrave, Xenical, Alli: Before Index Date",
                           "Qsymia, Contrave, Xenical, Alli: On or After Index Date",
                           "Qsymia, Contrave, Xenical, Alli Duration",
                           "Semaglutide (Ozempic)",
                           "Semaglutide (Ozempic) Duration",
                           "Semaglutide (Wegovy)",
                           "Semaglutide (Wegovy) Duration",
                           "Tirzepatide (Mounjaro)",
                           "Tirzepatide (Mounjaro) Duration",
                           "Albiglutide (Tanzeum)",
                           "Albiglutide (Tanzeum) Duration")
```

We will make a new column to consolidate which GLP-1 RA was issued at index date.

```{r}
# Define the GLP-1 RA medication columns
glp1_columns <- c("Dulaglutide (Trulicity)",
                  "Exenatide (Bydureon, Byetta)",
                  "Liraglutide (Saxenda)",
                  "Liraglutide (Victoza)",
                  "Lixisenatide (Adlyxin)",
                  "Semaglutide (Ozempic)",
                  "Semaglutide (Wegovy)",
                  "Tirzepatide (Mounjaro)",
                  "Albiglutide (Tanzeum)")

# Create the new column
summary_data$`GLP-1RA at Index Date` <- colnames(summary_data[, glp1_columns])[apply(summary_data[, glp1_columns], 1, which.max)]

# View the first few rows to check the result
head(summary_data$`GLP-1RA at Index Date`)
```

We will create a column to consolidate the GLP-1 RA duration as well.

```{r}
# Define the GLP-1RA duration columns
glp1_duration_columns <- c("Dulaglutide (Trulicity) Duration",
                           "Exenatide (Bydureon, Byetta) Duration",
                           "Liraglutide (Saxenda) Duration",
                           "Liraglutide (Victoza) Duration",
                           "Lixisenatide (Adlyxin) Duration",
                           "Semaglutide (Ozempic) Duration",
                           "Semaglutide (Wegovy) Duration",
                           "Tirzepatide (Mounjaro) Duration",
                           "Albiglutide (Tanzeum) Duration")

# Create a new column that sums the durations across all GLP-1RA drugs
summary_data$`GLP-1RA Duration` <- rowSums(summary_data[, glp1_duration_columns], na.rm = TRUE)

# View the first few rows to check the result
head(summary_data$`GLP-1RA Duration`)

```

## Creating a summary table with patient characteristics

`tbl_summary` was used to create a summary table.

```{r}
# Create the summary table including the new GLP-1RA at Index Date column
summary_table <- summary_data %>%
  select(Gender, `Age at Index Date`, `GLP-1RA at Index Date`) %>%
  tbl_summary(
    label = list(
      Gender = "Gender",
      `Age at Index Date` = "Age at Index Date",
      `GLP-1RA at Index Date` = "GLP-1 RA medication at Index Date"
    ),
    statistic = list(all_categorical() ~ "{n} ({p}%)"),  # Show counts and percentages
    missing_text = "Missing"
  ) %>%
  bold_labels()  # Bold the labels

# Display the summary table
summary_table

```

The summary table can be saved down if required.

```{r}
# Convert summary_table to a data frame (assuming it's a tbl_summary object)
df_summary <- summary_table %>% 
  as.data.frame()

# Save as CSV
write.csv(df_summary, "./data/cohort_summary_table_example.csv", row.names = FALSE)
```

## Creating a treatment duration summary table

For this, we will include other anti-obesity medications as well.

```{r}
# Identify columns containing "Duration"
duration_columns <- c("Qsymia, Contrave, Xenical, Alli Duration",
                      "Dulaglutide (Trulicity) Duration",
                       "Exenatide (Bydureon, Byetta) Duration",
                       "Liraglutide (Saxenda) Duration",
                       "Liraglutide (Victoza) Duration",
                       "Lixisenatide (Adlyxin) Duration",
                       "Semaglutide (Ozempic) Duration",
                       "Semaglutide (Wegovy) Duration",
                       "Tirzepatide (Mounjaro) Duration",
                       "Albiglutide (Tanzeum) Duration",
                      "GLP-1RA Duration")

# Select only the duration columns
duration_data <- summary_data %>%
  select(all_of(duration_columns))

```

We will work out the mean and standard deviation.

```{r}
# Calculate mean and standard deviation for duration columns
duration_stats <- duration_data %>%
  summarise(across(everything(), list(mean = ~ mean(.x, na.rm = FALSE),
                                      sd = ~ sd(.x, na.rm = FALSE))))
```

Next, the summary table is created.

```{r}
# Create a summary table for duration columns
duration_summary_table <- duration_data %>%
  tbl_summary(
    statistic = list(
      all_continuous() ~ "{mean} ({sd})"  # Mean and SD for continuous duration variables
    ),
    missing_text = ""
  ) %>% 
  # Remove rows with missing data (if necessary)
  remove_row_type(type = "missing", level_value = "Missing")

# Modify the table body to remove "Duration" from characteristics
duration_summary_table <- duration_summary_table %>%
  modify_table_body(~ {
    .x %>%
      mutate(across(everything(), ~ gsub("Duration", "", .)))
  }) %>%
  modify_header(label = '**Mean Prescription Duration**')

# Update the summary table header to "Mean Prescription Duration"
duration_summary_table <- duration_summary_table %>%
  modify_header(
    list(
      all_continuous() ~ "**Mean Prescription Duration**"
    )
  )

# Display the updated summary table
duration_summary_table
```

Again, the result can be saved.

```{r}
# Convert summary_table to a data frame (assuming it's a tbl_summary object)
df_dur_summary <- duration_summary_table %>% 
  as.data.frame()

# Save as CSV
write.csv(df_dur_summary, "./data/cohort_summary_table2.csv", row.names = FALSE, quote = TRUE)
```

## Stratifying the patient characteristics by clusters

We will use the cluster file saved down in notebook 05.

```{r}
# file directory for cluster information
clusters_file <- "./results/PatId_2-clusters_Model_A_stratified.csv"

# read the data
cluster_details <- read.table(clusters_file, header = TRUE, sep = ",", row.names = 1, check.names = FALSE)

# retrieve number of clusters
k <- length(unique(cluster_details$Cluster))

# update the values to be prefixed by "Cluster"
cluster_details <- cluster_details %>%
  mutate(across(Cluster, ~ factor(., levels = c(1:k), labels = paste("Cluster", 1:k))))

# preview
head(cluster_details)
```

Next, we will merge with the summary data.

```{r}
# Merge the data frames by row names
clustered_summary_data <- merge(summary_data, cluster_details, by = "row.names")

# Rename the 'Row.names' column back to row names if needed
rownames(clustered_summary_data) <- clustered_summary_data$Row.names
clustered_summary_data$Row.names <- NULL

# View the merged data
head(clustered_summary_data)
```

```{r}
# Stratify the summary table by Cluster
summary_table_stratified <- clustered_summary_data %>%
  select(Gender, `Age at Index Date`, `Qsymia, Contrave, Xenical, Alli: Before Index Date`,
         `Qsymia, Contrave, Xenical, Alli: On or After Index Date`, `GLP-1RA at Index Date`, Cluster) %>%
  tbl_summary(
    by = Cluster,  # Stratify by Cluster
    label = list(
      Gender = "Gender",
      `Age at Index Date` = "Age at Index Date",
      `GLP-1RA at Index Date` = "GLP-1 RA medication at Index Date"
    ),
    statistic = list(all_categorical() ~ "{n} ({p}%)"),  # Show counts and percentages
    missing_text = "Missing"
  ) %>%
  add_p() %>%  # Add p-values
  bold_labels()  # Bold the labels

# Display the stratified summary table
summary_table_stratified
```

Again, this can be saved down.

```{r}
# Convert stratified summary table to a data frame
df_summary_stratified <- summary_table_stratified %>% 
  as.data.frame()

# Save as CSV with updated filename
summary_strat_filename <- sprintf("./data/cohort_summary_table_stratified_by_cluster_%s.csv", "model_A_stratified")
write.csv(df_summary_stratified, summary_strat_filename, row.names = FALSE)
```

## Stratifying treatment durations by clusters

```{r}
# Convert duration columns to numeric, if they aren't already
clustered_summary_data[duration_columns] <- lapply(clustered_summary_data[duration_columns], as.numeric)
```

We will drop the medications which have no data.

```{r}
# Define columns to drop
columns_to_drop <- c("Lixisenatide (Adlyxin) Duration", "Albiglutide (Tanzeum) Duration")

# Drop the specified columns from the dataframe
filtered_summary_data <- clustered_summary_data %>%
  select(-all_of(columns_to_drop))

# Verify the columns have been dropped
print(colnames(filtered_summary_data))
```

```{r}
# Remove duration columns you don’t want
remaining_duration_columns <- setdiff(duration_columns, columns_to_drop)

# Filter data to include only the remaining duration columns
filtered_duration_data <- filtered_summary_data %>%
  select(all_of(remaining_duration_columns), Cluster)
str(filtered_duration_data)
```

```{r}
# Create the summary table for duration columns with filtered data
duration_summary_table_stratified_filtered <- filtered_duration_data %>%
  tbl_summary(
    by = Cluster,  # Stratify by Cluster
    statistic = list(
      all_continuous() ~ "{mean} ({sd})"  # Mean and SD for continuous duration variables
    ),
    missing_text = ""
  ) %>% 
  # Remove rows with missing data (if necessary)
  remove_row_type(type = "missing") %>% 
  modify_table_body(~ {
    .x %>%
      mutate(across(everything(), ~ gsub("Duration", "", .)))
  }) %>%
  modify_header(label = '**Mean Prescription Duration**')

# Update the summary table header to "Mean Prescription Duration"
duration_summary_table_stratified_filtered <- duration_summary_table_stratified_filtered %>%
  modify_header(
    label =  "**Mean Prescription Duration**"
    )

# Display the updated duration summary table
duration_summary_table_stratified_filtered
```

This can be saved as a csv file.

```{r}
# Convert the duration summary table to a data frame
df_dur_summary_stratified_filtered <- duration_summary_table_stratified_filtered %>% 
  as.data.frame()

# Save as CSV with updated filename
str_filtered_dur_filename <- sprintf("./data/cohort_duration_summary_table_stratified_by_cluster_filtered_%s.csv", model_name)
write.csv(df_dur_summary_stratified_filtered, str_filtered_dur_filename, row.names = FALSE, quote = TRUE)

```
